Macroscopic Salt Rejection through Electrostatically Gated Nanoporous Graphene

### 2019-08-20 paper/combined.tex
Change after final decision from NL

{\begin{flushleft}
\textbf{Teaser}:\\
The quantum capacitance-induced surface potential can stop electrolyte transport through a "large" graphene nanopore.
\end{flushleft}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Main text
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage{}
\section*{Introduction}
\label{sec:intro}
A central objective in the emerging field of nanofluidics is to
control transport phenomena at nanometer scale to explore new
technological opportunities. The reduction of characteristic
dimensions in nanofluidics gives rise to substantially different
transport properties in comparison to their bulk counterparts
\cite{Schoch_2008}. In nanopores, for example, the effect of
surface-mediated transport becomes dominant when the pore size is
comparable to the characteristic length of diffusion, enabling new
applications such as ionic diodes \cite{Karnik_2007,siwy2002fabrication,vlassiouk2007nanofluidic}, field-effect
transistors \cite{Nam_2009}, desalination \cite{Heiranian_2015}, and
nanopore-based DNA sequencing \cite{Heerema_2016,Garaj_2013}. Porous,
atomically thin two-dimensional (2D) materials, such as graphene, are
emerging as ideal membrane materials due to the ultimate permeation
\cite{Suk_2010,Jiang_2009,Celebi_2014,Koenig_2012,Drahushuk_2012}. Early
research has suggested that using a sheet of nanoporous graphene can
act as a desalination membrane by having the pore sizes smaller than
the hydrated radii ($\sim$1 nm), thereby hindering ion passage without
losing water permeability
\cite{Cohen_Tanugi_2012,Suk_2014,Cohen_Tanugi_2014,Cohen_Tanugi_2015,O_Hern_2014,O_Hern_2015,Surwade_2015,Walker_2017,Ghosh_2018}. Nevertheless,
in practice, the fabrication of subnanometer porous 2D membranes over
a large area with atomic precision remains technically challenging, 
{limiting the effective separation or demanding leak-sealing mechanism such as interfacial polymerization}
\cite{Suk_2014,Rollings_2016,O_Hern_2012,Wang_2017}. Moreover, a small
portion of excessively large pores or tears can significantly diminish
salt rejection.  Very recently, a number of reports have explored the
effect of surface charges on the selectivity of ion transport through
graphene nanopores, as a consequence of the electrostatic interactions
between solvated ions and the charged functional groups attached on
graphene, indicating that even large pores ($>$ 30nm) can be ion
selective \cite{Rollings_2016,Surwade_2014}. However, it remains
controversial that if the surface charges can result in a degree of
salt rejection, as the surface charge and potential cannot be
precisely quantified and are sensitive to the surroundings as well as
the history of chemical treatment \cite{Li_2008}. There remains a lack
of fundamental understanding of the interactions between the charged
2D nanopores and the electrolyte solution, { especially in regard of controlled ion transport modulation}.

In this report, we
investigate electrolyte permeation through graphene nanopores that are
significantly larger than the hydrated radii. We model the surface
potential by taking into account of the elementary electronic
properties of graphene and demonstrate that a high degree of salt
rejection can be achieved by electrostatic gating.

@@

Fig. \ref{fig:1}a depicts the experimental setup characterizing
ionic diffusion through a sheet of double layer porous graphene
(PG). Two reservoirs, namely the high-concentration reservoir (HCR)
containing electrolyte solution with molar concentration $c_0$, and
the low-concentration reservoir (LCR) containing de-ionized (DI)
water, are separated by a PG membrane supported by a polycarbonate
track-etched (PCTE) film. Magnetic stirrers are used to minimize the
effects of external concentration polarization, and a conductivity
probe is placed in the LCR to monitor the ionic concentration as a
function of time. The membrane is attached to a piece of copper tape
connected to a voltage source applying an electrical bias
$V_{\mathrm{G}}$, with the LCR grounded (Fig. \ref{fig:1}a
inset). We fabricated the PCTE-supported PG as schematically shown in
Fig. \ref{fig:1}b, using the protocol developed by our group
\cite{Choi_2018} (details see Supplementary Section
\ref{sec:exp}). The normally distributed, circular pores with pore
diameters of 20$\pm$10 nm and a pore density of
1.25$\pm$0.25$\times$10$^{10}$ cm$^{-2}$ (Fig. \ref{fig:1}c inset)
over a large area were perforated on graphene.  By optimizing the
subsequent transfer process, a high surface coverage ($>$ 98\%) of PG
on PCTE was obtained (Fig. \ref{fig:1}c), allowing reliable
characterization of the ion transport behavior (details see
Supplementary Section \ref{sec:exp}).  The system presented here
allows us to systematically investigate the diffusive flux of ions
across a sheet of PG, $\boldsymbol{J}$, driven by the gradient of
electrochemical potential in the electrolyte medium.

@@

Using 0.1 mM KCl solution in HCR, we obtained $\xi$ as a function of
$V_{\mathrm{G}}$ within $\pm1.25$ V, before triggering the
electrochemical reactions (Fig. \ref{fig:2}c). We observe an
asymmetric dependence of $\xi$ with respect to $V_{\mathrm{G}}$, with
a higher degree of salt rejection for $V_{\mathrm{G}}>0$. where
positive carriers (holes) are induced in graphene. We notice that an
increase of $\xi$, or a decrease of diffusive flux through PG with
$V_{\mathrm{G}}$, showing an inverse trend compared with those
observed in ionic field effect transistors (IFETs) in which the
diffusive flux increases with the gate voltage
\cite{Nam_2009,Cheng_2018}. A further increase of the KCl
concentration $c_{0}$ influences the attainable degree of salt
rejection. 
{
We note that the small negative $\xi$ values (corresponding to reduced transport resistance) at $V_{\mathrm{G}} \sim{}$ -0.5 V
may be caused by (i) the charge neutral point (CNP) of graphene is not exactly at $V_{\mathrm{G}}$ = 0, and (ii) experimental uncertainty.
Nevetherless, significant salt rejection is observed when $V_{\mathrm{G}}>$ 0.5 V.
}
The $\xi$ values measured at $V_{\mathrm{G}}$ = 1.25 V,
$\xi_{\mathrm{max}}$, as a function of $c_{0}$ from 10$^{-4}$ to
10$^{-2}$ M KCl (see Supplementary Section \ref{sec:debye}) exhibit
a power law dependency (Fig. \ref{fig:2}d). The
$\xi_{\mathrm{max}}-c_{0}$ relation can be nicely fitted by a power
law function following
$\xi_{\mathrm{max}} c_{0}^{0.52} = \mathrm{constant}$. As the Debye
length in solution, $\lambda_{\mathrm{D}}$, is inversely proportional
to $c^{0.5}$, we infer that the $V_{\mathrm{G}}$-induced salt
rejection originates from the modulation of the electrical double
layer (EDL), as will be discussed later.

@@

Therefore we adapt the simplification that the charge neutrality point

@@

velocity of graphene (more details find Supplementary Section
\ref{sec:charge-dens}).  Note that the PG film fabricated
experimentally is double-layer turbostratic (randomly aligned)
graphene, in which we assume the

@@

Eq. \eqref{eq:delta-phiG}. The theory proposed here was solved

@@

\ref{fig:trap}), which is related to the fact that observed $\xi$
is much lower within the $V_{\mathrm{G}}<0$ regime in Fig.

@@

example, Fig. \ref{fig:3}c presents the calculated

@@

Supplementary Section \ref{fig:scheme}). Indeed, the interplay
between graphene quantum capacitance and the EDL significantly reduces
the graphene surface potential $\psi_{\mathrm{G}}$ to $\sim$0.3 V at

@@

graphene with other nanoporous conducting 2D materials, although the
operational gate voltage window required to achieve the same degree of
salt rejection may be narrower, because graphene has the smallest
quantum capacitance (near the Dirac point) among all known 2D
materials \cite{Tian_2016}.

@@

regime, due to (i) the drift flux become dominant over diffusion flux

@@

are based on the setup  of ionic transport through single nanopore.

@@

$x_{r_{i}}$ is the fraction of pores with radius
$r_{i}$ (details see Supplementary Section
\ref{sec:pore-dist}). Combining Eq. \ref{eq:rejection-comb} and
Fig. \ref{fig:5}b, the
calculated salt rejection is reduced to $\sim{}$ 80\% of that from the
single-pore calculation of a 20 nm nanopore when $c_{0}$ = 0.1 mM (Supplementary Fig.
\ref{fig:simple-rect-pore}). Consider the existence of large pores
with diameter up to 60 nm, such degree of salt rejection is still
appreciable, which is ascribed to the nonlinear dependency of $\xi$ on
$\lambda_{\mathrm{D}} / r_{\mathrm{G}}$ and $V_{\mathrm{G}}$ (details
see Supplementary Section \ref{sec:pore-dist}).}

@@

The objective of the circuit analog model is to find effective salt
rejection ratio through PG, $\xi$ extracted from the experimental
$\eta$ values (defined in main text Eq. \ref{eq:rejection}). By
defining $R_{\mathrm{G}}^{0}=R_{\mathrm{G}}(V_{\mathrm{G}}=0)$,
$\delta = R_{\mathrm{PCTE}} / R_{\mathrm{G}}^{0}$,
$\chi = R_{\mathrm{G}}/R_{\mathrm{G}}^{0}$, $\eta$ is rewritten using
transport resistances as:
\begin{equation}
  \begin{aligned}[t]
      \label{eq:eta-R}
  \eta &= 1 - {\displaystyle
    \frac{(R_{\mathrm{PCTE}} + R_{\mathrm{G}})^{-1}}
    {(R_{\mathrm{PCTE}} + R_{\mathrm{G}}^{0})}} \\
    &= 1 - \frac{\delta + 1}{\delta + R_{\mathrm{G}}/R_{\mathrm{G}}^{0}} \\
    &= 1 - \frac{\delta + 1}{\delta + \chi}
  \end{aligned}
\end{equation}
Consequentially, the
salt rejection ratio $\xi$, is associated with $\chi$ by
$\xi = 1 - \chi^{-1}$ and is related with $\delta$ and $\eta$ via:
\begin{equation}
  \label{eq:xi-deriv}
  \begin{aligned}[t]
    \xi &= 1 - \chi^{-1} \\
    &= \frac{(\delta + 1) \eta}{\delta\eta + 1}
  \end{aligned}
\end{equation}
which is main text Eq. \ref{eq:xi-def}. When $0<\eta<1$ (salt
rejection regime) and $\delta>0$, Eq. \ref{eq:xi-deriv} always
guarantees $\xi>\eta$, since $\delta + 1$ is always larger than
$\delta\eta + 1$. In other words, due to the existence of
$R_{\mathrm{PCTE}}$, the effective salt rejection through PG upon
gating is always larger than that experimentally observed.

@@

\section{Measurement of charge carrier density of patterned graphene}
\label{sec:charge-dens}
The intrinsic charge carrier density of graphene was measured using a
field electron transistor (FET) in air with a 300nm SiO$_2$ as a
gate. The device setup is shown in Fig. \ref{fig:charge-dens}a,
where the source and drain electrodes were deposited onto the PG
sample transferred onto 300nm SiO$_2$ on doped Silicon. The
characteristic drain-source current $I_{\mathrm{DS}}$ as a function of
$V_{\mathrm{G}}$ is shown in Fig. \ref{fig:charge-dens}. The
charge-neutral point gate voltage $V_{\mathrm{CNP}}$ is measured to be 2.8 V.
Using the capacitance of the SiO$_{2}$ dielectric, we an intrinsic
charge density of $\sim{}$1.4$\times$10$^{11}$
\textit{e}$\cdot$cm$^{-2}$, and a shift in the PG's Fermi level by
$\sim{}$48 meV, which is negligible compared with the gate-induced
charge density in our experiments. Therefore, the assumption that the
PG sample is intrinsic without gating, is justified.

>>>>

{\begin{flushleft}
\textbf{Teaser}:\\
The quantum capacitance-induced surface potential can stop electrolyte transport through a ``large'' graphene nanopore.
\end{flushleft}
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Main text
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\clearpage{}
\section*{Introduction}
\label{sec:intro}
A central objective in the emerging field of nanofluidics is to
control transport phenomena at nanometer scale to explore new
technological opportunities. The reduction of characteristic
dimensions in nanofluidics gives rise to substantially different
transport properties in comparison to their bulk counterparts
\cite{Schoch_2008}. In nanopores, for example, the effect of
surface-mediated transport becomes dominant when the pore size is
comparable to the characteristic length of diffusion, enabling new
applications such as ionic diodes \cite{Karnik_2007,siwy2002fabrication,vlassiouk2007nanofluidic}, field-effect
transistors \cite{Nam_2009}, desalination \cite{Heiranian_2015}, and
nanopore-based DNA sequencing \cite{Heerema_2016,Garaj_2013}. Porous,
atomically thin two-dimensional (2D) materials, such as graphene, are
emerging as ideal membrane materials due to the ultimate permeation
\cite{Suk_2010,Jiang_2009,Celebi_2014,Koenig_2012,Drahushuk_2012}. Early
research has suggested that using a sheet of nanoporous graphene can
act as a desalination membrane by having the pore sizes smaller than
the hydrated radii ($\sim$1 nm), thereby hindering ion passage without
losing water permeability
\cite{Cohen_Tanugi_2012,Suk_2014,Cohen_Tanugi_2014,Cohen_Tanugi_2015,O_Hern_2014,O_Hern_2015,Surwade_2015,Walker_2017,Ghosh_2018}. Nevertheless,
in practice, the fabrication of subnanometer porous 2D membranes over
a large area with atomic precision remains technically challenging, 
limiting the effective separation
  % or demanding leak-sealing mechanism such as interfacial polymerization
\cite{Suk_2014,Rollings_2016,O_Hern_2012,Wang_2017}. Moreover, a small
portion of excessively large pores or tears can significantly diminish
salt rejection. Regarding these issues, we have previously developed a
highly-scalable technique to fabricate uniformly-distributed nanopores
(sub-20 to 50 nm) over wafer-scale graphene samples
\cite{Choi_2018}. A major step towards its real-world desalination and
separation applications, would be to develop methodology that allow
appreciable selectivity of nanoporous 2D membrane with pore size even
larger than the hydration radii. A number of recent reports have
explored the effect of surface charges on the selectivity of ion
transport through graphene nanopores, as a consequence of the
electrostatic interactions between solvated ions and the charged
functional groups attached on graphene, indicating that even large
pores ($>$ 30nm) can be ion selective
\cite{Rollings_2016,Surwade_2014}. However, it remains controversial
that if the surface charges can result in a degree of salt rejection,
as the surface charge and potential cannot be precisely quantified and
are sensitive to the surroundings as well as the history of chemical
treatment \cite{Li_2008}. There remains a lack of fundamental
understanding of the interactions between the charged 2D nanopores and
the electrolyte solution, { especially in regard of controlled ion
  transport modulation}.

In this report, we investigate electrolyte permeation through graphene
nanopores that are significantly larger than the hydration radii. We
model the surface potential by taking into account the elementary
electronic properties of graphene and demonstrate that a high degree
of salt rejection could be achieved by both pore sizing and
electrostatic gating.

@@

Fig. \ref{fig:1}a depicts the experimental setup characterizing
ionic diffusion through a sheet of double layer porous graphene
(PG). Two reservoirs, namely the high-concentration reservoir (HCR)
containing electrolyte solution with molar concentration $c_0$, and
the low-concentration reservoir (LCR) containing de-ionized (DI)
water, are separated by a PG membrane supported by a polycarbonate
track-etched (PCTE) film. Magnetic stirrers are used to minimize the
effects of external concentration polarization, and a conductivity
probe is placed in the LCR to monitor the ionic concentration as a
function of time. The membrane is attached to a piece of copper tape
connected to a voltage source applying an electrical bias
$V_{\mathrm{G}}$, with the LCR grounded (Fig. \ref{fig:1}a
inset). We fabricated the PCTE-supported PG as schematically shown in
Fig. \ref{fig:1}b, using the protocol developed by our group
\cite{Choi_2018}, which  enables formation of uniformly-distributed graphene  nanopores over large area up
 to 25 cm$^{2}$ (details see Supplementary Section
\ref{sec:exp}). The normally distributed, circular pores with pore
diameters of 20$\pm$10 nm and a pore density of
1.25$\pm$0.25$\times$10$^{10}$ cm$^{-2}$ (Fig. \ref{fig:1}c inset)
over a large area were perforated on graphene.  By optimizing the
subsequent transfer process, a high surface coverage ($>$ 98\%) of PG
on PCTE was obtained (Fig. \ref{fig:1}c), allowing reliable
characterization of the ion transport behavior (details see
Supplementary Section \ref{sec:exp}).  The system presented here
allows us to systematically investigate the diffusive flux of ions
across a sheet of PG, $\boldsymbol{J}$, driven by the gradient of
electrochemical potential in the electrolyte medium.

@@

Using 0.1 mM KCl solution in HCR, we obtained $\xi$ as a function of
$V_{\mathrm{G}}$ within $\pm1.25$ V, before triggering the
electrochemical reactions (Fig. \ref{fig:2}c). We observe an
asymmetric dependence of $\xi$ with respect to $V_{\mathrm{G}}$, with
a higher degree of salt rejection for $V_{\mathrm{G}}>0$. where
positive carriers (holes) are induced in graphene. We notice that an
increase of $\xi$, or a decrease of diffusive flux through PG with
$V_{\mathrm{G}}$, shows an inverse trend compared with those
observed in ionic field effect transistors (IFETs) in which the
diffusive flux increases with the gate voltage
\cite{Nam_2009,Cheng_2018}. A further increase of the KCl
concentration $c_{0}$ influences the obtainable degree of salt
rejection. 
We note small negative $\xi$ values (corresponding to reduced transport resistance) at $V_{\mathrm{G}} \sim{}$ -0.5 V
likely caused by (i) offset of the charge neutral point (CNP) of graphene is from $V_{\mathrm{G}}$ = 0 and (ii) experimental uncertainty.
Nevetherless, significant salt rejection is observed when $V_{\mathrm{G}}>$ 0.5 V.
The $\xi$ values measured at $V_{\mathrm{G}}$ = 1.25 V,
$\xi_{\mathrm{max}}$, as a function of $c_{0}$ from 10$^{-4}$ to
10$^{-2}$ M KCl (see Supplementary Section \ref{sec:debye}) exhibit
a power law dependency (Fig. \ref{fig:2}d). The
$\xi_{\mathrm{max}}-c_{0}$ relation can be nicely fitted by a power
law function following
$\xi_{\mathrm{max}} c_{0}^{0.52} = \mathrm{constant}$. As the Debye
length in solution, $\lambda_{\mathrm{D}}$, is inversely proportional
to $c^{0.5}$, we infer that the $V_{\mathrm{G}}$-induced salt
rejection originates from the modulation of the electrical double
layer (EDL), as will be discussed later.

@@

Therefore, we adopt the simplification that the charge neutrality point

@@

velocity of graphene \cite{Tian_2016}.  Note that the PG film
fabricated experimentally is double-layer turbostratic (randomly
aligned) graphene, in which we assume the

@@

Eq. \eqref{eq:delta-phiG}. The influence of polymer residue on the
surface of graphene is not considered in the current theory due to:
(i) insulating polymer residues do not respond to the gate voltage and
(ii) the thickness of the polymer residue is minimal after the thermal
annealing process. The theory proposed here was solved

@@

\ref{fig:trap}), which is related to the fact that observed $\xi$ is
much lower within the $V_{\mathrm{G}}<0$ regime in Fig.

@@

example, Fig. \ref{fig:3}b presents the calculated

@@

Supplementary Section \ref{sec:numer}). Indeed, the interplay between
graphene quantum capacitance and the EDL significantly reduces the
graphene surface potential $\psi_{\mathrm{G}}$ to $\sim$0.3 V at

@@

graphene with other nanoporous conducting 2D materials, while they are
less practical due to the narrower operational window of gate voltage
required to achieve the same level of salt rejection. This is because
graphene has the smallest quantum capacitance (near the Dirac point)
among all known 2D materials \cite{Tian_2016}.

@@

regime, due to scenarios that (i) the drift flux become dominant over diffusion flux

@@

are based on the setup  of ionic transport through a single nanopore.

@@

$x_{r_{i}}$ is the fraction of pores with radius $r_{i}$ (details see
Supplementary Section \ref{sec:pore-dist}). Combining
Eq. \ref{eq:rejection-comb} and Fig. \ref{fig:5}b, the calculated salt
rejection is reduced to $\sim{}$ 80\% of that from the single-pore
calculation of a 20 nm nanopore when $c_{0}$ = 0.1 mM (Supplementary
Fig.  \ref{fig:simple-rect-pore}). Consider the existence of large
pores with diameter up to 60 nm, such degree of salt rejection is
still appreciable, which is ascribed to the nonlinear dependency of
$\xi$ on $\lambda_{\mathrm{D}} / r_{\mathrm{G}}$ and $V_{\mathrm{G}}$
(details see Supplementary Section \ref{sec:pore-dist}).}  From our
analysis, it is also clear that reducing the average pore size is critical
to overcome the low salt rejection at higher concentrations, which may
hopefully be achieved with relevant fabrication techniques.

@@

The objective of the circuit analog model is to find effective salt
rejection ratio through PG, $\xi$ extracted from the experimental
$\eta$ values (defined in main text Eq. \ref{eq:rejection}). By
defining $R_{\mathrm{G}}^{0}=R_{\mathrm{G}}(V_{\mathrm{G}}=0)$,
$\delta = R_{\mathrm{PCTE}} / R_{\mathrm{G}}^{0}$,
$\chi = R_{\mathrm{G}}/R_{\mathrm{G}}^{0}$, $\eta$ is rewritten using
transport resistances as:
\begin{equation}
  \begin{aligned}[t]
      \label{eq:eta-R}
  \eta &= 1 - {\displaystyle
    \left[\frac{(R_{\mathrm{PCTE}} + R_{\mathrm{G}})}
    {(R_{\mathrm{PCTE}} + R_{\mathrm{G}}^{0})}\right]^{-1}} \\
    &= 1 - \frac{\delta + 1}{\delta + R_{\mathrm{G}}/R_{\mathrm{G}}^{0}} \\
    &= 1 - \frac{\delta + 1}{\delta + \chi}
  \end{aligned}
\end{equation}
Consequentially, the
salt rejection ratio $\xi$, is associated with $\chi$ by
$\xi = 1 - \chi^{-1}$ and is related with $\delta$ and $\eta$ via:
\begin{equation}
  \label{eq:xi-deriv}
  \begin{aligned}[t]
    \xi &= 1 - \chi^{-1} \\
    &= \frac{(\delta + 1) \eta}{\delta\eta + 1}
  \end{aligned}
\end{equation}
which is main text Eq. \ref{eq:xi-def}. When $0<\eta<1$ (salt
rejection regime) and $\delta>0$, Eq. \ref{eq:xi-deriv} always
guarantees $\xi>\eta$, since $\delta + 1$ is always larger than
$\delta\eta + 1$. In other words, due to the existence of
$R_{\mathrm{PCTE}}$, the effective salt rejection through PG upon
gating is always larger than that experimentally observed.

@@

\section{Measurement of charge carrier density of patterned graphene}
\label{sec:charge-dens}
The intrinsic charge carrier density of graphene was measured using a
field electron transistor (FET) in air with a 300nm SiO$_2$ as a
gate. The device setup is shown in Fig. \ref{fig:charge-dens}a,
where the source and drain electrodes were deposited onto the PG
sample transferred onto 300nm SiO$_2$ on doped Silicon. The
characteristic drain-source current $I_{\mathrm{DS}}$ as a function of
$V_{\mathrm{G}}$ is shown in Fig. \ref{fig:charge-dens}. The
charge-neutral point gate voltage $V_{\mathrm{CNP}}$ is measured to be 2.8 V.
Using the capacitance of the SiO$_{2}$ dielectric, we estimate an intrinsic
charge density of $\sim{}$1.4$\times$10$^{11}$
\textit{e}$\cdot$cm$^{-2}$, and a shift in the PG's Fermi level by
$\sim{}$48 meV, which is negligible compared with the gate-induced
charge density in our experiments. Therefore, the assumption that the
PG sample is intrinsic without gating, is justified.
